---
title: "Reproducible Research: Peer Assessment 1 - Barbara Rattner"
output: 
  html_document:
    keep_md: true
---
---
title: "Assignment 1 - Reproducible Research Course (week 2)"
author: "Barbara Rattner"
date: "10/1/2019"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

The data are loaded from a csv file in the working directory called "activity.csv"

```{r}
activity.data <- read.csv("activity.csv")
```

## What is the mean total number of steps taken per day?

*(For this part of the assignment, I am ignoring missing values)*

1. Calculate the total number of steps taken per day

To calculate the steps taken per day, we need to first calculate the daily number of steps.

```{r}
StepsPerDay <- tapply(activity.data$steps, activity.data$date, sum, na.rm = TRUE)
```
To present the data in table format I will use the 'kable' function in the 'knitr' package.

```{r}
library(knitr)
kable(StepsPerDay, caption = "Steps taken per day", col.names ="Steps")
```

2. Make a histogram of the total number of steps taken each day

```{r}
hist(StepsPerDay, col = "gray", xlab = "Steps per Day", breaks = 15, main = paste("Frequency of" , "total number of steps taken in a day")) 
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
Mean <- format(mean(StepsPerDay), nsmall = 2)
Median <- format(median(StepsPerDay), nsmall = 2)
```

The **mean** number of steps taken per day is **`r Mean`** and the **median** number of steps taken per day is **`r Median`**.


## What is the average daily activity pattern?

1. Make a time series plot (type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

``` {r}
AverageStepsPerInterval <- aggregate(x = list(AverageSteps = activity.data$steps), by = list(Interval = activity.data$interval), FUN=mean, na.rm = TRUE)
 plot(AverageStepsPerInterval, type = "l", main = "Average Number of Steps Taken Daily in Interval", xlab = "5-Minute Interval", ylab = "Average Number of Steps per Day")
```




2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?




## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
